tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-15 paper.tex
WIP

\section{Results}
\subsection{Phenotypic Model}
The additive model is a common assumption in GWAS
\citep{uffelmann2021}, and we assume that a quantitative trait $y$
is given by
\[
y=X\beta+\epsilon,
\]
where $\beta$ is the effect size, $X$ is a matrix describing the number
of causal alleles in the individual, and $\epsilon$ is the environmental
noise.
[TODO more discussion here of what this means in an ARG. Be clear about
causal mutations]
% Environmental noise is normally distributed with mean zero
% and a variance of $V_G(1-h^2)/{h^2}$,
% % distribution,
% % \begin{align}\label{eq:env}
%     % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
% % \end{align}
% where $V_G$ is the additive genetic variance from the simulated effect sizes,
% and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

@@

\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is assocated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
and an effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of their corresponding node values.
\label{fig:tree-illustration}}
\end{figure}

>>>>

\section{Results}
\begin{figure}[t]%
\centering
\includegraphics[width=240pt]{figures/tree-illustration.pdf}
\caption{Example simulation of a phenotype at a site with ancestral state A and
two mutations. In this diploid example each of the three individuals
is assocated with two nodes (i.e., the individual with ID 0 corresponds
to nodes \textsf{a} and \textsf{b}). Internal nodes in the tree are associated
with the null individual, $-1$. Here, the trait's causal allele is T
and effect size $\beta=0.05$. Each node in the tree has an associated
genetic value, and the overall genetic value for an individual is the
sum of their corresponding node values. The final phenotype
for each individual is the sum of the genetic value and
simulated environmental noise.
\label{fig:tree-illustration}}
\end{figure}

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
Each trait is associated with one or more causal sites (position
on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide). Because we are reasoning
about the inheritance of particular alleles at a given position,
we can consider just the local tree at that site and
the network structure of the ARG along the genome is not
important~\citep{wong2023general}. In
Fig.~\ref{fig:tree-illustration} we show the local tree
relating three diploid individuals for a given causal site.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour). Ancestral
nodes are not associated with individuals.

[WIP]
At a given site, the causal allele is associated with an
effect size $\beta$,
and every node in the tree that carries this allele has a
``genetic value'' of $\beta$.

Following the standard practise in GWAS~\citep{uffelmann2021}
we assume the additive model, and a quantitative trait $y$
is given by
\[
y=X\beta+\epsilon,
\]
where $\beta$ is the effect size, $X$ is a matrix describing the number
of causal alleles in the individual, and $\epsilon$ is the environmental
noise.

Environmental noise is normally distributed with mean zero
and a variance of $V_G(1-h^2)/{h^2}$,
% distribution,
% \begin{align}\label{eq:env}
    % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
% \end{align}
where $V_G$ is the additive genetic variance from the simulated effect sizes,
and $h^2$ is the narrow-sense heritability provided as input by the user.
The \texttt{tstrait} API is modular, however, and other forms of environmental
noise can be easily generated by the user, if they wish.

@@
